Motmot: A Distributed Chat Protocol

### 2012-05-04 paper/paper.tex
Most of design

This simplification to Paxos yields many benefits.  For instance, clients may
still choose to keep a record of chats in a local file, but by disallowing
protocol rejoins, this persistent storage does not play any role in the safety
or progress of \libmotmot's Paxos implementation.  We can also avoid the need
for complicated and high-bandwidth recovery mechanisms.  In \libmotmot's Paxos,
as in the final protocol presented in the original Paxos paper, only one
participant is designated as the proposer, and hence allowed to issue prepares
and subsequently decrees, at any given time.  In the absence of rejoins, we
are able to provide a very useful system invariant which full Paxos cannot
have:
\begin{quote}
  If a proposer successfully completes a prepare phase, that proposer will
  remain the protocol's designated proposer until it leaves the protocol, by
  failure or otherwise.
\end{quote}
We discuss this invariant further in the following subsection.

@@

\begin{description}
  \item[Join, part, and election:]
    The original Paxos paper remarks on, but does not actually describe, a
    procedure for electing a Paxos system's designated proposer; nor does it
    describe mechanisms for adding and removing individuals from the system.
    These are all, however, necessary to the functionality of our consensus
    algorithm.  Election is necessary in order to ensure that the system makes
    progress past the prepare phase of Paxos, lest multiple participants
    attempt repeatedly to prepare over one another.  Join and part are required
    in order to support any usable notion of chat.

    In \libmotmot, we guarantee the consistency of joins, parts, and election
    results by baking them in to the standard commit protocol.

  \item[Greet protocol:]

  \item[Out-of-band requests:]

@@

  \item[Commit retry:]

  \item[Log synchronization and truncation:]

>>>>

This simplification to Paxos yields many benefits.  For instance, clients may
still choose to keep a record of chats in a local file, but by disallowing
protocol rejoins, this persistent storage does not play any role in the safety
or progress of \libmotmot's Paxos implementation.  We can also avoid the need
for complicated and high-volume recovery mechanisms.  In \libmotmot's Paxos,
as in the final protocol presented in the original Paxos paper, only one
participant is designated as the proposer, and hence allowed to issue prepares
and subsequently decrees, at any given time.  In the absence of rejoins, we
are able to provide a very useful system invariant which full Paxos cannot
have:
\begin{quote}
  If a proposer successfully completes a prepare phase, that proposer will
  remain the protocol's designated proposer until it leaves the protocol, by
  failure or otherwise.
\end{quote}
We discuss this invariant further in the following subsection.

@@

\begin{description}
  \item[Join, part, and election:]
    The original Paxos paper mentions, but does not actually provide, a
    procedure for electing a Paxos system's designated proposer; nor does it
    describe mechanisms for adding and removing individuals from the system.
    These are all, however, necessary to the functionality of our consensus
    algorithm.  Election is necessary in order to ensure that the system makes
    progress past the prepare phase of Paxos, lest multiple participants
    attempt repeatedly to prepare over one another.  Join and part are required
    in order to support any usable notion of chat.

    In \libmotmot, we guarantee the consistency of joins, parts, and election
    results by baking them in to the standard commit protocol.  Each protocol
    participant keeps a list of all the participants in the chat, ordered
    by session-unique identifiers.  Participants can be added and removed from
    this list only through Paxos commits decreeing either \verb`DEC_JOIN` or
    \verb`DEC_PART`, which ensures the consistency of the participants in this
    list and their order.  Participants are identified by the ID of the Multi-%
    Paxos instance which decreed their join.

    Proposer election is performed locally, with each participant simply
    recognizing the earliest-joined participant to whom they have a live socket
    connection as the designated proposer.  Participants will only accept a
    prepare phase issued by this designated proposer.  When a new proposer
    successfully completes its prepare phase, it broadcasts a decree requesting
    that all participants evict the old proposer from their participant lists.
    In this manner, we are able to guarantee the invariant described above.

    This mechanism may occasionally result in the forced eviction of a live
    participant who was the previous proposer.  It is conceivable that transient
    network problems might cause a majority of participants to lose their
    connection to the original proposer, and hence to accept the prepare of
    the new proposer.  Before the new proposer is able to issue its decree
    parting the old proposer, however, these connections may all be
    reestablished, and the old proposer may be booted even though the system
    as a whole currently recognizes its presence.  This causes no loss of
    consistency and merely forces an old proposer in this situation to rejoin
    the chat.  We consider this a worthwhile tradeoff given the usefulness
    of our invariant.

  \item[Greet protocol:]
    The presence of joins and parts demands a mechanism for initiating
    communication with new protocol participants.  When a join is successfully
    accepted by a majority, the proposer will decree it.  Subsequently, the
    proposer will send the new participant a welcome message, which transfers
    all of its protocol state to the new participant.  The proposer always has
    the most up-to-date state, and since the network guarantees FIFO message
    ordering and the proposer is the only one who can issue commands to modify
    protocol state, the new participant's state will be consistent.  The
    proposer also assigns the new participant its ID (i.e., the instance
    number of its join decree).

    The new participant will then open sockets to each participant on the
    list it receives from the proposer.  It will then send each participant
    a hello message.  This hello message is also used in the reconnect protocol
    and is used to bind sockets to the protocol state objects representing
    participants.

  \item[Out-of-band requests:]
    The proposer is the only protocol participant able to order and to decree
    a chat message.  Hence, all participants must have a mechanism for asking
    the proposer to decree a message.  These requests also apply for joins
    (though not to parts, since we do not support forced kicking of users from
    a chat).

    We want to avoid stressing the network, and in particular want to keep the
    volume of data the proposer must handle to a minimum, since the proposer is
    one of the system's primary bottlenecks.  Thus, we did not want to force
    the proposer to broadcast the arbitrarily-large chat messages produced by
    the various users in the chat.  Instead of sending along the actual message,
    the requester of a message generates a session-unique identifier for the
    message, consisting of the requester's own participant ID and a locally
    unique request ID.  The requester then takes on the responsibility of
    broadcasting this identifier along with the actual message contents to all
    participants.  When this message is received, each participant caches it
    locally.  The proposer will make a decree which contains the identifier of
    the request; when the other participants receive a commit for the decree,
    they can do a local lookup for the message contents.

    This has the added benefit of providing all participants an assurance that
    message contents come directly from their sources without the possibility
    of corruption by the proposer.  At worst, the proposer can modify the
    cache index of the message.

    Because our request mechanism is done out-of-band, commits for a given
    request may arrive after the message contents are received from the initial
    broadcast.  We implement a basic protocol for retrieving and resending
    these messages if they cannot be found.

@@

  \item[In-order learn:]
    \libmotmot's clients are notified of Paxos commits via a set of callbacks.
    In order to ensure the total ordering of chats as seen by the user, we
    ``learn'' about commits in order of Paxos instance number.  This means that
    a missing commit message could prevent client-side progress.  To rectify
    this, we implement a straightforward means for any participant to ask the
    proposer to remind them of a committed value.

  \item[Log synchronization and truncation:]
    Commits made in Paxos must be kept by all protocol participants in order
    to obtain Paxos's correctness guarantees.  Since we keep all of our logs
    in memory, however, we need a mechanism for pruning information which
    the entire system has finished learning.

    We implement an out-of-band sync-and-truncate protocol, which is used by
    the proposer to issue consistent system-wide deletion of old commits.  A
    proposer asks all protocol participants to return the last contiguous
    commit which their respective clients have learned about.  Once it receives
    a response from \emph{all} participants, it can take the minimum instance
    number of those commits and safely command all participants to truncate
    up to that point.
